---
title: "SERI Fire Seminar - Initial Scenarios"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

RHESSys simulations were conducted on a 1 km^2^, mixed conifer watershed in the southern Sierra Nevada. The watershed has an elevation ranging from 1790 to 2115 m and is located at the rain/snow transition zone, with a snow fraction ranging from 35% to 60% of total precipitation. The watershed receives approximately 1320 mm of precipitation in an average year, mostly during the winter season. 

Vegetation was modeled as having two canopies, an upper canopy representing conifers and a lower canopy representing understory vegetation (aka ladder fuels), which may be a mix of young trees, shrubs and other low stature vegetation. Each simulation began with vegetation growing from bare ground, which is analogous to starting the simulations after a stand-replacing fire. Although the model is distibuted and simulates ecohydrologic processes for different areas of the watershed, our analysis is restricted to basin level averages.

Each simulation runs for 80-years, from 1941 to 2021. (The observed meteorogical data actually runs from 1941 to 2005, but the record repeats again starting in 2006). A moderate intensity wildfire occurs after 25 years (1966) and a high intensity wildfire occurs 35 years later in 2001. In the first, or baseline, senario, the watershed is only affected by the two wildfires. In the second scenario, a small fuels treatment is implemented in 1956 and 30% of the understory biomass is removed.

**Scenario #1 (Baseline):**  
* Start: Oct 1, 1941  
* Fire: Aug 1, 1966  
* Fire: Aug 1, 2001  
* End: Sept 30, 2021  

**Scenario #2 (Single fuel treatment):**  
* Start: Oct 1, 1941  
* Fuel treatment: Aug 1, 1956 (30% understory removal)  
* Fire: Aug 1, 1966  
* Fire: Aug 1, 2001  
* End: Sept 30, 2021  

## Model Outputs

Each scenario in RHESSys produces a timeseries of output for a number of variables. These variables are provided in two text files. The file basin_results.txt contains all variables expect for canopy specific variables which are found in canopy_results.txt.    

**basin_results.txt**  
* date  
* year  
* month  
* day  
* wy - wateryear (Oct 1 to Sept 30)  
* wyd - wateryear day  
* scenario  
* precip - Precipitation (mm)  
* tmax - Max temperature (C)  
* tmin - Min temperature (C)  
* streamflow (mm)  
* snowpack (mm)  
* pet - Potential evapotranspiration (mm)  
* et - Actual evapotranspiration (mm)  
* photosyn - Photosynthesis (gC/m2)  
* litter (gC/m2)  

The file canopy_results.txt contains variables that the vegetation canopies. This file has an additional variable called canopy_layer that differentiates the upper and lower canopy for the watershed.

**canopy_results.txt**  
* date  
* year  
* month  
* day  
* wy - wateryear (Oct 1 to Sept 30)  
* wyd - wateryear day  
* scenario  
* canopy_layer - Upper canopy (1) or Lower Canopy (2)  
* Height (m)  
* Above-ground Carbon (gC/m2)  
* Below-ground Carbon (gC/m2)  

## Potential Additional Scenarios

There are a number of options for selecting additional scenarios for addressing your research question. Changes that are available include:  
* Modifying the climate sequence (precipitation and/or temperature)  
* Change the type of fuels treatments  (for example, overstory vs understory)  
* Change the timing of fuels treatments  
* Change the size of fuels treatments  


## Analyze scenarios in R

```{r libraries}
library(ggplot2)
library(dplyr)
```

Import data. (Note that you will need to change the path to your data)

``` {r import}
basin_results <- read.table("../outputs/basin_results.txt", header = T)
canopy_results <- read.table("../outputs/canopy_results.txt", header = T)
basin_results$date <- as.Date(basin_results$date)
canopy_results$date <- as.Date(canopy_results$date)
```

Let's plot how streamflow differs between the two scenarios. Note that two scenarios are colored differently.

``` {r streamflow}
ggplot(basin_results) +
  geom_line(aes(x=date, y=streamflow, color=as.factor(scenario)))
```

There is a lot of data, which makes it hard to evaluate. Let's group the data by wateryear.

``` {r streamflow_wy}
basin_results_groups <- dplyr::group_by(basin_results, wy, scenario)
basin_results_wy <- dplyr::summarise(basin_results_groups,
                                     streamflow_wy = sum(streamflow))
ggplot(basin_results_wy) +
  geom_line(aes(x=wy, y=streamflow_wy, color=as.factor(scenario)))
```

We can use a boxplot to easier compare annual streamflow between the two scenarios.

``` {r boxplot_wy}
basin_results_groups <- dplyr::group_by(basin_results, wy, scenario)
basin_results_wy <- dplyr::summarise(basin_results_groups,
                                     streamflow_wy = sum(streamflow))
ggplot(basin_results_wy) +
  geom_boxplot(aes(x=wy ,y=streamflow_wy, color=as.factor(scenario)))
```

Now lets see an example of simply plotting the first scenario instead of both scenarios.

``` {r streamflow_wy_scenario_1}
basin_results_scenario_1 <- dplyr::filter(basin_results, scenario==1)
basin_results_groups <- dplyr::group_by(basin_results_scenario_1, wy, scenario)
basin_results_wy <- dplyr::summarise(basin_results_groups,
                                     streamflow_wy = sum(streamflow))
ggplot(basin_results_wy) +
  geom_line(aes(x=wy, y=streamflow_wy, color=as.factor(scenario)))
```


Now lets plot above-ground carbon for both scenarios. Since there are two canopies, we have to differentiate between the upper and lower canopy, in addition to scenarios.  

``` {r above_ground_carbon}
ggplot(canopy_results) +
  geom_line(aes(x=date, y=above_ground_carbon, color=as.factor(scenario), linetype=as.factor(canopy_layer)))
```

And once-again, grouped by wateryear.

``` {r above_ground_carbon_wy}
canopy_results_groups <- dplyr::group_by(canopy_results, wy, scenario, canopy_layer)
canopy_results_wy <- dplyr::summarise(canopy_results_groups,
                                      above_ground_carbon_wy = mean(above_ground_carbon))
ggplot(canopy_results_wy) +
  geom_line(aes(x=wy, y=above_ground_carbon_wy, color=as.factor(scenario), linetype=as.factor(canopy_layer)))

```


